---
title: "Lab5"
author: "Jeffrey Kinard"
date: "11/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(ggplot2)
library(knitr)
library(readxl)
library(tidyverse)
```

```{r}
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
pers <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

**1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?**  
  
The below barchart clearly shows that the weekend has much more accidents than during the week. This is possibly due to the fact that people are more inclined to travel and shop on weekends since they don't have work. With this extra congestion in shopping centers, and people traveling to new areas, accidents are more likely to occur.

```{r}
summary <- acc %>% select(c(DAY_WEEK)) %>% group_by(DAY_WEEK) %>% summarise(num=n())
summary$days = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")
summary$days = factor(c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"), levels=unique(summary$days))
summary %>% ggplot(aes(x=days, y=num)) + geom_bar(stat="identity")
```

**2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)**

```{r}
severe <- pers %>% filter(INJ_SEV==4)
head(severe)
```

**3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)**

```{r}
danger <- pers %>% 
  filter(INJ_SEV==4) %>% 
  select(STATE, MAKE) %>% 
  group_by(STATE, MAKE) %>% 
  summarise(t=n()) %>% 
  drop_na() %>% 
  group_by(STATE) %>% 
  filter(t==max(t)) %>% 
  select(STATE, MAKE)
danger
```

**4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)**  
  
The below map shows the most dangerous vehicle manufacturer in each state. The most dangerous vehicle is the vehicle that causes the most fatalities in traffic accidents.

```{r}
glc <- readxl::read_xlsx("/Users/jeff/Downloads/FRPP_GLC_-_United_StatesSep292020.xlsx")
new_glc <- glc %>% 
  rename(STATE="State Code") %>% 
  mutate(STATE=as.numeric(STATE))
new_danger <- danger %>% 
  inner_join(danger %>% 
               ungroup() %>% 
               select(MAKE) %>% 
               unique() %>% 
               mutate(NAME=c("Ford", "Chevrolet", "Toyota", "Honda")), by="MAKE") %>% 
  inner_join(new_glc %>% 
               select(STATE, `State Name`) %>% 
               unique(), by="STATE")
states <- map_data("state")
new_danger$`State Name` <- tolower(new_danger$`State Name`)
map <- new_danger %>% 
  rename(region=`State Name`) %>% 
  left_join(states, by="region") 
makeNames <- map %>% group_by(region, NAME) %>% summarize(long=mean(long), lat=mean(lat)) %>% drop_na()
map %>% 
  ggplot(aes(x = long, y = lat)) +
  geom_polygon(aes(group=group), fill="white", color="black") +
  geom_text(aes(label=NAME), color='black', data=makeNames)
```
  
Since the above map can be hard to read, the below map shows the same data, but each state is color-coded by make of car.

```{r}
map %>% 
  ggplot(aes(x = long, y = lat, fill=NAME)) +
  geom_polygon(aes(group=group))
```

**5. Join the accident and person table (work out which variable(s) to use)**

```{r}
acc_pers <- acc %>% inner_join(pers, by=c("STATE", "ST_CASE", "VE_FORMS", "COUNTY", "DAY", "MONTH", "HOUR", "MINUTE", "RUR_URB", "FUNC_SYS", "HARM_EV", "MAN_COLL", "SCH_BUS"))
head(acc_pers)
```

**6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.**  
  
The below barcharts show the number of accidents by hour in the day per gender for each day of the week. There are more men involved in traffic accidents than women for every data point collected.

```{r}
day <- data.frame(days=c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"), DAY_WEEK=c(1, 2, 3, 4, 5, 6, 7))
day$days = factor(c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"), levels=unique(day$days))
sex = data.frame(gender=c("Male", "Female"), SEX=c(1, 2))

info <- acc_pers %>% select(SEX, DAY_WEEK, HOUR) %>% filter((SEX==1 | SEX==2) & HOUR < 99) %>% group_by(SEX, DAY_WEEK, HOUR) %>% summarise(total=n()) %>% inner_join(day, by="DAY_WEEK") %>% inner_join(sex, by="SEX")
info %>% ggplot(aes(x=HOUR, y=total, fill=gender)) + geom_bar(position="dodge", stat="identity") + facet_wrap(~days)
```

**7. Now plot a choropleth map of the number of deaths on a county level.**  
  
The below map shows the number of deaths that occur in traffic accidents by county in the US. The areas most affected are the LA region of California and the Houston area of Texas.

```{r}
fatal <- acc %>% select(STATE, COUNTY, FATALS) %>% group_by(STATE, COUNTY) %>% summarise(deaths=sum(FATALS))
county <- map_data("county")
fatal_glc <- glc %>% 
  rename(STATE=`State Code`, COUNTY=`County Code`) %>% 
  mutate(STATE=as.numeric(STATE), COUNTY=as.numeric(COUNTY))

countyMap <- fatal %>% 
  ungroup() %>% 
  inner_join(fatal_glc, by=c("STATE", "COUNTY")) %>% 
  select(deaths, `State Name`, `County Name`) %>% 
  unique() %>%
  rename(region=`State Name`, subregion=`County Name`) %>% 
  mutate(region=tolower(region), subregion=tolower(subregion))%>% 
  right_join(county, by=c("region", "subregion"))

countyMap$deaths[is.na(countyMap$deaths)] <- 0

countyMap %>% ggplot(aes(x = long, y = lat, fill=deaths)) +
  geom_polygon(aes(group=group))
```

**8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.**  
  
The below maps show the sum of deaths caused by traffic accidents per state. The maps compare the winter and summer months to see if there is a change in how many accidents occur. It is suprising to see that summer is actually more dangerous in terms of traffic accidents since winter typically has more dangerous weather conditions. This could be due to a change in driving behavior that occurs in winter. Drivers take more caution in winter months which helps prevent accidents. In the summer months, drivers take less caution and that causes more mistakes.

```{r}
summer <- acc %>% 
  filter(MONTH>=6 | MONTH<=8) %>% 
  select(STATE, FATALS) %>% 
  group_by(STATE) %>% 
  summarise(summer=sum(FATALS))
winter <- acc %>% 
  filter(MONTH==12 | MONTH<=2) %>% 
  select(STATE, FATALS) %>% 
  group_by(STATE) %>% 
  summarise(winter=sum(FATALS))
seasons <- summer %>% 
  inner_join(winter, by="STATE") %>% 
  pivot_longer(!STATE, names_to="season", values_to="deaths")

states <- map_data("state")
states_glc <- glc %>% 
  rename(STATE=`State Code`) %>% 
  mutate(STATE=as.numeric(STATE))

stateMap <- seasons %>% 
  ungroup() %>% 
  left_join(states_glc, by="STATE") %>% 
  select(deaths, season, `State Name`) %>% 
  unique() %>%
  rename(region=`State Name`) %>% 
  mutate(region=tolower(region))%>% 
  right_join(states, by="region")

stateMap %>% ggplot(aes(x = long, y = lat, fill=deaths)) +
  geom_polygon(aes(group=group)) +
  facet_grid(~season)
```